A Ph.D. can be a solitary endeavor, even more so when it is interrupted by a pandemic. I owe a great debt to the people who have helped remedy that. This page will unfortunately only have an incomplete list of them.

First and foremost, I want to thank my adviser, Andrew W. Appel, for his patience with me, and his foresight and wisdom in research. I gained great insight and an eclectic taste of research from all the times he answered my questions by referring me to different papers he wrote in the 1990s. He also gave me all the time and space I needed to dip my toes in different fields of computer science and quench my curiosity with many side projects, some of which have even led to papers. He provided extensive feedback on countless iterations of this dissertation, for which I am deeply grateful. 

I would like to thank the members of my thesis committee, Zachary Kincaid and Matthieu Sozeau, for taking the time to read my dissertation and to give me feedback, and Aarti Gupta and David Walker, for their support and questions.

I would like to thank my colleague, Kathrin Stark, with whom I have discussed many of the ideas presented in this thesis. Without her contributions, the concepts I present in my thesis would be much less refined. I would like to thank Tim Carstens for the time he spent collaborating with us. His contributions to the garbage collector and his use of our project have illuminated our path forward.

I would also like to thank Anastasiya Kravchuk-Kirilyuk, Joseph W. Cutler, and the programming languages group at Princeton, for their academic and social support. I would like to thank Julien Vanegue for encouraging me to wrap up my thesis and also making it easier for me to do so.

I would like to thank my parents, Hülya and Birol Korkut, for providing me with all the resources I needed since I was a child with a burning interest in computers. This dissertation would not have been possible without their sacrifice and their belief in me.

Finally, I would like to thank my wife, Anna Blech, for the love and joy she brings to my life, and for her unwavering support throughout graduate school and beyond. Without her feedback, this dissertation would have been full of unnecessary commas, run-on sentences, and uninspired prose.

\newpage

This material is based upon work supported by the National Science Foundation under Grant No.\ CCF-2005545. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author and do not necessarily reflect the views of the National Science Foundation.